The \dealii Library, Version 9.7

### 2025-07-30 9.7/paper.tex
Update the release date.
Add a note on PSBLAS.
Mention the new anisotropic simplex conversions.

\dealii version 9.7.0 was released August 11, 2024.
\todo{Update}
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii software library version 9.7. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
terms of the \emph{GNU Lesser General Public License} (LGPL). The \dealii
project is in the process of relicensing the library under the terms of
the \emph{Apache License 2.0 with LLVM Exception}. Downloads are
available at \url{https://www.dealii.org/} and
\url{https://github.com/dealii/dealii}.

@@

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
\item MPGMRES and GMRES updates
  \todo[inline]{Matthias and Martin, please update (or make this point
  into one or two subsections in Section 2 below).}
\item A common operation in the finite element codes is to ask whether
  a specific shape function belongs to a particular solution
  variable. In the past, this was implemented by using the
  \texttt{FiniteElement::system\_to\_component\_index()} function that
  returns, among other information, which vector component a shape
  function belongs to. One then had to test whether this is the vector
  component (or one of the vector components) that corresponds to a
  specific variable. The new
  \texttt{FiniteElement::shape\_function\_belongs\_to()} makes this
  easier: It takes as argument a \texttt{FiniteElementExtractor} that
  corresponds to a scalar, vector-valued, or tensor-valued set of
  solution variables, and directly returns \texttt{true} or
  \texttt{false}.
\item \todo[inline]{Add to this list other points that are not
    explicitly discussed below}
\end{itemize}
%
The
\href{https://dealii.org/current/doxygen/deal.II/changes_between_9_6_0_and_9_7_0.html}{changelog}
-- listing more than 120
features and bugfixes --
contains a complete record of all changes; see \cite{changes97}.

@@

\todo[inline]{Perhaps talk about this point:
  GridGenerator::convert\_hypercube\_to\_simplex\_mesh() can now split
  quadrilaterals into two triangles and hexahedra into six
  tetrahedra. These anisotropic splits are useful in contexts where it
  is important to add as few new elements as possible.}

@@

>>>>

\dealii version 9.7.0 was released July 22, 2025.
\todo{Update}
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii software library version 9.7. \dealii is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
terms of the \emph{GNU Lesser General Public License} (LGPL). The \dealii
project is in the process of relicensing the library under the terms of
the \emph{Apache License 2.0 with LLVM Exception}. Downloads are
available at \url{https://www.dealii.org/} and
\url{https://github.com/dealii/dealii}.

@@

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
\item MPGMRES and GMRES updates
  \todo[inline]{Matthias and Martin, please update (or make this point
  into one or two subsections in Section 2 below).}
\item A common operation in the finite element codes is to ask whether
  a specific shape function belongs to a particular solution
  variable. In the past, this was implemented by using the
  \texttt{FiniteElement::system\_to\_component\_index()} function that
  returns, among other information, which vector component a shape
  function belongs to. One then had to test whether this is the vector
  component (or one of the vector components) that corresponds to a
  specific variable. The new
  \texttt{FiniteElement::shape\_function\_belongs\_to()} makes this
  easier: It takes as argument a \texttt{FiniteElementExtractor} that
  corresponds to a scalar, vector-valued, or tensor-valued set of
  solution variables, and directly returns \texttt{true} or
  \texttt{false}.
\item Most of the functions in the \texttt{GridGenerator} namespace correspond
  to simple geometries (such as cubes and spheres) discretized with hypercubes.
  The function \texttt{GridGenerator::convert\_hypercube\_to\_simplex\_mesh()}
  converts such discretizations to instead use simplices. That function now
  supports anisotropic splits, including splitting quadrilaterals into two
  triangles and hexahedra into six tetrahedra, in addition to the standard
  isotropic splits. These anisotropic splits significantly reduce the total
  number of cells and are useful in contexts where it is important to add as few
  new cells as possible.

\item \todo[inline]{Add to this list other points that are not
    explicitly discussed below}
\end{itemize}
%
The
\href{https://dealii.org/current/doxygen/deal.II/changes_between_9_6_0_and_9_7_0.html}{changelog}
-- listing more than 120
features and bugfixes --
contains a complete record of all changes; see \cite{changes97}.

@@

\todo[inline]{I moved the `improved anisotropic split' change listed here to the
  `other changes' section.}

@@

\todo[inline]{Luca, want to say something here? If not, please delete
  the subsubsection.}
\todo[inline]{DRW: note that while we support configuring with PSBLAS we don't
  actually implement any interfaces to it yet (i.e., we have one test but no
  code in the library which uses \texttt{DEAL\_II\_WITH\_PSBLAS}).}
